### 2019-03-12 2-Chapters/6-Chapter/nonstatbandits.git/NonStatB.tex
Almost done reordering chapter 6 and cutting useless parts. I guess the chapter will be finished tomorrow.

\TODOL{This chapter is basically a raw include from my paper ``The Generalized Likelihood Ratio Test meets klUCB: an Improved Algorithm for Piece-Wise Non-Stationary Bandits'', see https://hal.inria.fr/hal-02006471}

We propose a new algorithm for the piece-wise \iid{} non-stationary bandit problem with bounded rewards.
Our proposal, \GLRklUCB, combines an efficient bandit algorithm, \klUCB, with an efficient, \emph{parameter-free}, change-point detector, the Bernoulli Generalized Likelihood Ratio Test, for which we provide new theoretical guarantees of independent interest. We analyze two variants of our strategy, based on \emph{local restarts} and \emph{global restarts}, and show that their regret is upper-bounded by  $\cO(\Upsilon_T \sqrt{T \log(T)})$ if the number of change-points $\Upsilon_T$ is unknown, and by $\cO(\sqrt{\Upsilon_T T \log(T)})$ if $\Upsilon_T$ is known.
This improves the state-of-the-art bounds, as our algorithm needs no tuning based on knowledge of the problem complexity other than $\Upsilon_T$.
We present numerical experiments showing that \GLRklUCB{} outperforms passively and actively adaptive algorithms from the literature, and highlight the benefit of using local restarts.

% -----------------------------------------------------------------
\section{Motivation for Non-Stationary MAB Models}
\label{sec:6:Introduction}

Multi-Armed Bandit (MAB) problems form a well-studied class of sequential decision making problems, in which an agent repeatedly chooses an action $I_t \in\{1,\dots,K\}$ or ``arm'' --in reference to the arm of a one-armed bandit-- among a set of $K$ arms. In the most standard version of the stochastic bandit model, each arm $i$ is associated with an \iid{} sequence of rewards $(r_{i,t})$ that follow some distribution of mean $\mu_i$. Upon selecting arm $I_t$, the agent receives the reward $r_{I_t,t}$ associated to the chosen arm, and her goal is to adopt a sequential sampling strategy that maximize the expected sum of these rewards. This is equivalent to minimizing the \emph{regret}, defined as the difference between the total reward of the oracle strategy always selecting the arm with largest mean, $\mu^*$, and that of our strategy: $R_T = \bE[\sum_{t=1}^T(\mu^* - \mu_{I_t})]$.

Regret minimization in stochastic bandits has been extensively studied since the works of \cite{Robbins52} and \cite{LaiRobbins85}, and several algorithms with a $\cO(\log(T))$ \emph{problem-dependent} regret upper bound have been proposed (see, \eg, \cite{LattimoreBanditAlgorithmsBook} for a survey). Among those, the \klUCB{} algorithm (\cite{KLUCBJournal}) has been shown to be asymptotically optimal for Bernoulli distributions (in that it exactly matches the lower bound given by \cite{LaiRobbins85}) and can also be employed when the rewards are assumed to be bounded in $[0,1]$.
\emph{Problem-independent} upper bounds of the form $R_T=\Omega(\sqrt{KT})$ (with no hidden constant depending on the arms distributions) have also been established for stochastic algorithms, like MOSS, or \klUCB-Switch by \cite{Garivier18}, while \klUCB{} is known to enjoy a sub-optimal $\cO(\sqrt{KT\ln(T)})$ problem-independent regret.

Stochastic bandits were historically introduced as a simple model for clinical trials, where arms correspond to some treatments with unknown efficacy (\cite{Thompson33}). More recently, MAB models have been proved useful for different applications, like cognitive radio, where arms can model the vacancy of radio channels, or parameters of a dynamically configurable radio hardware (\cite{Maghsudi16,Bonnefoi17,KerkoucheAlami18}). Another application is the design of recommender systems, where arms model the popularity of different items (\eg, news recommendation, \cite{Li10}).

For both cognitive radio and recommender systems, the assumption that the arms distribution \emph{do not evolve over time} may be a big limitation. Indeed, in cognitive radio new devices can enter or leave the network, which impacts the availability of the radio channel they use to communicate; whereas in online recommendation, the popularity of items is also subject to trends. Hence, there has been some interest on how to take those \emph{non-stationary} aspects into account within a multi-armed bandit model.

@@

In this paper, we propose a new algorithm for the piece-wise stationary bandit problem with bounded rewards, called \GLRklUCB. Like previous approaches -- \CUSUM{} (\cite{LiuLeeShroff17}) and \MUCB{} (\cite{CaoZhenKvetonXie18}) -- our algorithm relies on combining a standard multi-armed bandit algorithm with a change-point detector. For the bandit component, we propose the use of the \klUCB{} algorithm, that is known to outperform {\UCB}1 (\cite{Auer02}) used in previous works. For the change-point detector, we propose the Bernoulli Generalized Likelihood Ratio Test (GLRT), for which we provide new non-asymptotic properties that are of independent interest. This choice is particularly appealing because unlike previous approaches, the Bernoulli GLRT is \emph{parameter-free}: it does not need the tuning of a window size ($w$ in \MUCB), or the knowledge of a lower bound on the magnitude of the smallest change ($\varepsilon$ in \CUSUM).

@@

The article is structured as follows. We introduce the model and review related works in Section~\ref{sec:6:BanditSetting}. In Section~\ref{sec:6:ChangePointDetector}, we study the Generalized Likelihood Ratio test (GLRT) as a Change-Point Detector (CPD) algorithm.
We introduce the two variants of \GLRklUCB{} algorithm in Section~\ref{sec:6:GLRklUCB_Algorithm}, where we also present upper bound on the regret of each variant. The unified regret analysis for these two algorithms is sketched in Section~\ref{sec:6:RegretAnalysis}. Numerical experiments
are presented in Section~\ref{sec:6:NumericalExperiments}, with more details in the Appendix.

% -----------------------------------------------------------------
\section{The Piece-Wise Stationary Bandit Setup and Related Works}
\label{sec:6:BanditSetting}

@@

A policy $\pi$ chooses the next arm to play based on the sequence of past plays and obtained rewards.
The performance of $\pi$ is measured by its (piece-wise stationary) \emph{regret}, the difference between the expected reward obtained by an oracle policy playing an optimal arm $i^*_t$ at time $t$, and that of the policy $\pi$:
\begin{align}\label{def:6:nti}
    R_T^{\pi} &= \E\left[\sum_{t=1}^T \left(\mu_{i^*_t}(t) - \mu_{I_t}(t)\right)\right].
\end{align}

In the piece-wise \iid{} model, we furthermore assume that there is a (relatively small) number of \emph{breakpoints}, denoted by $\Upsilon_T:=\sum_{t=1}^{T-1} \indic\left(\exists i\in\{1,\dots,K\}: \mu_t(i) \neq \mu_{t+1}(i)\right)$.
We define the $k$-th breakpoint by $\tau^{(k)} = \inf\{t > \tau^{(k-1)} : \exists i : \mu_i(t) \neq \mu_{i}(t+1)\}$. Hence for $t\in[\tau^{(k)} + 1,\tau^{(k+1)}]$, the rewards $(X_{i,t})$ associated to each arms are \iid.
Note than when a breakpoint occurs, we do not assume that all the arms means  change, but that \emph{there exists} an arm whose mean has changed. Depending on the application, many scenario can be meaningful: changes occurring on all arms simultaneously (due to some exogenous event), or only a few arms change at each breakpoint. Introducing the number of change-points  on arm $i$, defined as $\NCi:=\sum_{t=1}^{T-1} \indic\left(\mu_t(i) \neq \mu_{t+1}(i)\right)$, it clearly holds that
$\NCi \leq \Upsilon_T$, but there can be an arbitrary difference between these two quantities for some arms. Letting $C_T := \sum_{i=1}^K \NCi$ be the total number of change-points on the arms, one can have $C_T \in \{ \Upsilon_T, \dots, K\Upsilon_T \}$.

The piece-wise stationary bandit model can be viewed as an interpolation between stationary and adversarial models, as the stationary model corresponds to $\Upsilon_T = 0$, while the adversarial model can be considered as a special (worst) case, with $\Upsilon_T = T$. However, analyzing an algorithm for the piece-wise stationary bandit model requires to assume a small number of changes, typically $\Upsilon_T = o(\sqrt{T})$.

% -----------------------------------------------------------
\paragraph{Related work.}

We review previous works that studied the piece-wise stationary bandit model, or variants of this model.
To the authors knowledge, all the previous works are based on the idea of combining a classical bandit policy, such as Thompson sampling, \UCB{} or EXP3, with a strategy to account for changes in arms' distributions.
Following the vocabulary used in previous works, we make the distinction between passively and actively adaptive strategies.
Active strategies monitors the rewards of each arm by using a change detection algorithm (\cite{Basseville93}), and reset the history of pulls and rewards of one or all the arms as soon as a change is detected.
Passive strategies are tuned with a discount factor or a limited memory size, while active strategies are usually tuned with a threshold for the statistical test.

% \TODOL{Review of previous work (see section on Experiments), and state-of-the-art of passively adaptive and actively adaptive algorithms}

@@

Both D-UCB and SW-UCB builds on a stationary policy,
but for example EXP3.S from \cite{Auer02NonStochastic}
builds on the EXP3 policy, which is designed for the adversarial case,
and achieves a larger regret upper-bound.
We also found in our experiments that it always performed worse that the reference stationary algorithms (\eg, \klUCB), so we do not include it in Section~\ref{sec:6:NumericalExperiments}.
Similarly, previous works showed that older algorithms have no or weaker regret guarantees, and have been proved to be less efficient empirically, or are designed for more specific settings.
%
For instance, Windowed-Mean Shift from \cite{YuMannor09} is generic and combines any efficient bandit policy with a CD test based on a sliding window, but there approach is not applicable to the bandit setting as they consider side observations.
%
A different setting where the quantity of interest is not $\Upsilon_T$ but $V_T$ the total variational budget, was introduced by \cite{Besbes14stochastic}, for which they proposed the RExp3 algorithm.
%

The idea of using a simple discount factor, as for D-UCB, was recently adapted to a Bayesian policy, with the Discounted Thompson sampling (DTS) algorithm presented by \cite{RajKalyani17}.
Even if no theoretical guarantee was given, it can be empirically very efficient, but we found that DTS is not robust in the choice of if its discount factor $\gamma\in(0,1)$, contrarily to what was highlighted in their paper.
%
In general, we found that passively adaptive approaches can be efficient when their parameters are well tuned, but our experiments show that actively adaptive algorithms perform significantly better.

@@

When all their parameters are tuned correctly, both approaches are proved to achieve a good regret upper-bound.
As both results are valid under different assumptions, we consider the two results to be the current state-of-the-art.
As for previous works, tuning their parameters requires to know both the horizon $T$ and the number of breakpoints $\Upsilon_T$, but most importantly it requires a prior knowledge of the problem difficulty, by assuming to know a lower bound on both the time between changes (\eg, $L$ for \MUCB) and on the changes on the means of arms (\eg, $\varepsilon$ for \CUSUM).
%
\CUSUMUCB{} achieves a regret bounded by $\cO(\sqrt{\Upsilon_T T \log(T / \Upsilon_T)})$, but only for Bernoulli distributions,
and when its $4$ parameters ($\alpha,\varepsilon,M,h$) are tuned based on problem-dependent knowledge (see XXX).
%
\MUCB{} achieves a regret bounded by $\cO(\sqrt{\Upsilon_T T \log(T)})$, for bounded distributions, and when its $4$ parameters ($\alpha,w,b,\gamma$) are also tuned based on a problem-dependent knowledge of $\tilde{\delta}$ and $L$ (see Assumption~1 and Remark~1).

\TODOL{Finish presenting MUCB and CUSUM, and highlights the differences!}

\paragraph{Expert aggregation.}
%
The second line of research of actively adaptive algorithms uses a Bayesian point of view.
A Bayesian CD algorithm is combined with Thompson Sampling in \cite{MellorShapiro13},
and more recently \cite{Alami17} introduced the Memory Bandit algorithm. It is presented as efficient empirically, but no theoretical guarantee are given, and due to its complexity, we do not include it in our experiments.
The idea behind Memory Bandit is to use an expert aggregation algorithm, like EXP4 from \cite{Auer02}, modified to efficiently aggregate an growing number of experts from techniques presented in \cite{Mourtada17}.
At each time step, a new expert is introduced, and experts corresponds to different Thompson sampling algorithms, each using a different history of pulls and rewards. Intuitively, after a change-point the newest experts will soon become the most efficient, as they are learning by using rewards drawn from the new distribution(s).
%
Note that the regret bound for these methods are still open-problems.

@@

\TODOL{FIXME next paragraph is the short version of related works review I have to merge the long and short together and keep as much details as I can!}

The piece-wise stationary bandit model was first studied by \cite{Kocsis06,YuMannor09,Garivier11UCBDiscount}. It is also known as \emph{switching} (\cite{MellorShapiro13}) or \emph{abruptly changing stationary} (\cite{WeiSrivastava18}) environment. %We do not consider the \emph{slowly-varying} MAB models (\cite{Louedec16,WeiSrivastava18}), even if it is interesting for some applications.
% We review previous works that studied the piece-wise stationary bandit model.
To our knowledge, all the previous approaches combine a standard bandit algorithm, like \UCB{}, Thompson Sampling or EXP3, with a strategy to account for changes in the arms distributions. This strategy often consists in \emph{forgetting old rewards}, to efficiently focus on the most recent ones, more likely to be similar to future rewards. We make the distinction between \emph{passively} and \emph{actively} adaptive strategies.

The first proposed mechanisms to forget the past consist in either discounting rewards (at each round, when getting a new reward on an arm, past rewards are multiplied by $\gamma^n$ if that arm was not seen since $n>0$ times, for a discount factor $\gamma\in(0,1)$), or using a sliding window (only the rewards gathered in the $\tau$ last observations of an arm are taken into account, for a window size $\tau$).
Those strategies are passively adaptive as the discount factor or the window size are fixed, and can be tuned as a function of $T$ and $\Upsilon_T$ to achieve a certain regret bound.
Discounted UCB (D-UCB) was proposed by \cite{Kocsis06} and analyzed by \cite{Garivier11UCBDiscount}, who prove a $\cO(\sqrt{\Upsilon_T T }\log(T))$ regret bound, if $\gamma = 1 - \sqrt{\Upsilon_T/T}/4$.
The same authors proposed the Sliding-Window UCB (SW-UCB) and prove a $\cO(\sqrt{\Upsilon_T T \log(T)})$ regret bound, if $\tau = 2 \sqrt{T\log(T)/\Upsilon_T}$.

More recently, \cite{RajKalyani17} proposed the Discounted Thompson Sampling (DTS) algorithm, which performs well in practice with $\gamma=0.75$. However, no theoretical guarantees are given for this strategy, and our experiments did not really confirm the robustness to $\gamma$.
%
The RExp3 algorithm (\cite{Besbes14stochastic}) can also be qualified as passively adaptive: it is based on (non-adaptive) restarts of the EXP3 algorithm. Note that this algorithm is introduced for a different setting, where the quantity of interest is not $\Upsilon_T$ but a quantity $V_T$ called the total variational budget (satisfying $\Delta^{\text{change}} \Upsilon_T \leq V_T \leq \Upsilon_T$ with $\Delta^{\text{change}}$ the minimum magnitude of a change-point). A  $\cO(V_T^{1/3} T^{2/3})$ regret bound is proved, which is weaker than existing results in our setting. Hence we do no include this algorithm in our experiments.

% \TODOE{Il faut faire attention pour DTS, tu attaques leur papier! Qu'est-ce que nos expériences montrent exactement qui est en contradiction avec ce qu'ils disent?}
% \TODOE{Je vire EXP3.S: c'est une version d'EXP3 qui a des garanties en terme de ``nombre de breakpoint'' dans un cadre adversarial, pas stochastique. Il n'est pas basé sur des restart et un discount}
% \TODOE{Vérifie que je raconte pas de conneries pour RExp3, Odalric et Subho l'avaient mis dans la catégorie ``passive'' mais je comprends pas trop comment marche l'algo}

%
The first \emph{actively adaptive} strategy is Windowed-Mean Shift (\cite{YuMannor09}), which combines any bandit policy with a change point detector which performs \emph{adaptive restarts} of the bandit algorithm. However, this approach is not applicable to our setting as it takes into account side observations. Another line of research on actively adaptive algorithms uses a Bayesian point of view.
A Bayesian Change-Point Detection (CPD) algorithm is combined with Thompson Sampling by \cite{MellorShapiro13}, and more recently in the Memory Bandit algorithm of \cite{Alami17}. Both algorithms do not have theoretical guarantees and their implementation is very costly, hence we do not include them in our experiments. Our closest competitors rather use frequentist CPD algorithms (see, e.g. \cite{Basseville93}) combined with a bandit algorithm.
% The idea behind Memory Bandit is to use an expert aggregation algorithm, like EXP4 from \cite{Auer02}, modified to efficiently aggregate an growing number of experts from techniques presented in \cite{Mourtada17}.
% At each time step, a new expert is introduced, and experts correspond to different Thompson sampling algorithms, each using a different history of pulls and rewards. Intuitively, after a change-point the newest experts will soon become the most efficient, as they are learning by using rewards drawn from the new distribution(s).
%
The first algorithm of this flavor, Adapt-EVE algorithm (\cite{Hartland06}) uses a Page-Hinkley test and the \UCB{} policy, but no theoretical guarantee are given. EXP3.R (\cite{Allesiardo15,Allesiardo17}) combines a CPD with EXP3, and the history of all arms are reset as soon as a a sub-optimal arm is detecting to become optimal and it achieve a $\cO(\Upsilon_T \sqrt{T \log(T)})$ regret. This is weaker than the $\cO(\sqrt{\Upsilon_T T \log(T)})$ regret achieved by two recent algorithms, \CUSUMUCB{} (\cite{LiuLeeShroff17}) and Monitored UCB (\MUCB, \cite{CaoZhenKvetonXie18}).

\CUSUMUCB{} is based on a rather complicated two-sided \CUSUM{} test, that uses the first $M$ samples from one arm to compute an initial average, and then detects whether a drift of size larger than $\varepsilon$ occurred from this value by checking whether a random walk based on the remaining observations crosses a threshold $h$. Thus it requires the tuning of three parameters, $M$, $\varepsilon$ and $h$. \CUSUMUCB{} performs \emph{local restarts} using this test, to reset the history of \emph{one arm} for which the test detects a change.
\MUCB{} uses a much simpler test, based on the $w$ most recent observations from an arm: a change is detected if the absolute difference between the empirical means of the first and second halves of those $w$ observations exceeds a threshold $h$. So it requires the tuning of two parameters, $w$ and $h$. \MUCB{} performs \emph{global restarts} using this test, to reset the history of \emph{all arms} whenever the test detects a change on one of the arms.
% Both approaches also require a mechanism to add exploration, parameterized by $\alpha\in(0,1)$.
% \CUSUMUCB{} uses a randomized exploration, by sampling arm $i$ with probability $\alpha/K$ at each time step,
% and \MUCB{} uses a more deterministic scheme, to force exploring the $K$ arms (when the time since the last restart, $t - \tau$, is found to be in $\{1,\dots,K\}$ modulo $\lceil 1/\alpha \rceil$).
% Both solutions ensure that all arms are sampled enough on each stationary sequence, in order for the CD algorithm to have enough \iid{} samples to detect changes.
Compared to \CUSUMUCB{}, note that \MUCB{} is numerically much simpler as it only uses a bounded memory, of order $\bigO{K w}$ for $K$ arms.

% \TODOE{I chose not to talk about $\alpha$ here in the end to gain space: in the end, everyone share that parameter. Maybe add a footnote in section 4 saying that CUSUM does random exploration}

>>>>

\TODOL{Be much more concise, if I already give an abstract in the beginning of the chapter...}

% We propose a new algorithm for the piece-wise \iid{} non-stationary bandit problem with bounded rewards.
% Our proposal, \GLRklUCB, combines an efficient bandit algorithm, \klUCB, with an efficient, \emph{parameter-free}, change-point detector, the Bernoulli Generalized Likelihood Ratio Test, for which we provide new theoretical guarantees of independent interest. We analyze two variants of our strategy, based on \emph{local restarts} and \emph{global restarts}, and show that their regret is upper-bounded by  $\cO(\Upsilon_T \sqrt{T \log(T)})$ if the number of change-points $\Upsilon_T$ is unknown, and by $\cO(\sqrt{\Upsilon_T T \log(T)})$ if $\Upsilon_T$ is known.
% This improves the state-of-the-art bounds, as our algorithm needs no tuning based on knowledge of the problem complexity other than $\Upsilon_T$.
% We present numerical experiments showing that \GLRklUCB{} outperforms passively and actively adaptive algorithms from the literature, and highlight the benefit of using local restarts.

% -----------------------------------------------------------------
\section{Introduction}
\label{sec:6:Introduction}

\TODOL{Be much more concise: in the thesis, MAB models and applications have already been introduced, motivated etc!}

% Multi-Armed Bandit (MAB) problems form a well-studied class of sequential decision making problems, in which an agent repeatedly chooses an action $I_t \in\{1,\dots,K\}$ or ``arm'' --in reference to the arm of a one-armed bandit-- among a set of $K$ arms. In the most standard version of the stochastic bandit model, each arm $i$ is associated with an \iid{} sequence of rewards $(r_{i,t})$ that follow some distribution of mean $\mu_i$. Upon selecting arm $I_t$, the agent receives the reward $r_{I_t,t}$ associated to the chosen arm, and her goal is to adopt a sequential sampling strategy that maximize the expected sum of these rewards. This is equivalent to minimizing the \emph{regret}, defined as the difference between the total reward of the oracle strategy always selecting the arm with largest mean, $\mu^*$, and that of our strategy:

% \[ R_T = \bE[\sum_{t=1}^T(\mu^* - \mu_{I_t})]. \]

% Regret minimization in stochastic bandits has been extensively studied since the works of \cite{Robbins52} and \cite{LaiRobbins85}, and several algorithms with a $\cO(\log(T))$ \emph{problem-dependent} regret upper bound have been proposed (see, \eg, \cite{LattimoreBanditAlgorithmsBook} for a survey). Among those, the \klUCB{} algorithm (\cite{KLUCBJournal}) has been shown to be asymptotically optimal for Bernoulli distributions (in that it exactly matches the lower bound given by \cite{LaiRobbins85}) and can also be employed when the rewards are assumed to be bounded in $[0,1]$.
% \emph{Problem-independent} upper bounds of the form $R_T=\Omega(\sqrt{KT})$ (with no hidden constant depending on the arms distributions) have also been established for stochastic algorithms, like MOSS, or \klUCB-Switch by \cite{Garivier18}, while \klUCB{} is known to enjoy a sub-optimal $\cO(\sqrt{KT\ln(T)})$ problem-independent regret.

Stochastic bandits were historically introduced as a simple model for clinical trials, where arms correspond to some treatments with unknown efficacy (\cite{Thompson33}). More recently, MAB models have been proved useful for different applications, like cognitive radio, where arms can model the vacancy of radio channels, or parameters of a dynamically configurable radio hardware (\cite{Maghsudi16,Bonnefoi17,KerkoucheAlami18}). Another application is the design of recommender systems, where arms model the popularity of different items (\eg, news recommendation, \cite{Li10}).
%
For both cognitive radio and recommender systems, the assumption that the arms distribution \emph{do not evolve over time} may be a big limitation. Indeed, in cognitive radio new devices can enter or leave the network, which impacts the availability of the radio channel they use to communicate; whereas in online recommendation, the popularity of items is also subject to trends. Hence, there has been some interest on how to take those \emph{non-stationary} aspects into account within a multi-armed bandit model.

@@

In this chapter, we propose a new algorithm for the piece-wise stationary bandit problem with bounded rewards, called \GLRklUCB. Like previous approaches -- \CUSUM{} (\cite{LiuLeeShroff17}) and \MUCB{} (\cite{CaoZhenKvetonXie18}) -- our algorithm relies on combining a standard multi-armed bandit algorithm with a change-point detector. For the bandit component, we propose the use of the \klUCB{} algorithm, that is known to outperform {\UCB}1 (\cite{Auer02}) used in previous works. For the change-point detector, we propose the Bernoulli Generalized Likelihood Ratio Test (GLRT), for which we provide new non-asymptotic properties that are of independent interest. This choice is particularly appealing because unlike previous approaches, the Bernoulli GLRT is \emph{parameter-free}: it does not need the tuning of a window size ($w$ in \MUCB), or the knowledge of a lower bound on the magnitude of the smallest change ($\varepsilon$ in \CUSUM).

@@

\paragraph{Outline}
%
This chapter is structured as follows.
We introduce the model and review related works in Section~\ref{sec:6:BanditSetting}. In Section~\ref{sec:6:ChangePointDetector}, we study the Generalized Likelihood Ratio test (GLRT) as a Change-Point Detector (CPD) algorithm.
We introduce the two variants of \GLRklUCB{} algorithm in Section~\ref{sec:6:GLRklUCB_Algorithm}, where we also present upper bound on the regret of each variant. The unified regret analysis for these two algorithms is sketched in Section~\ref{sec:6:RegretAnalysis}. Numerical experiments
are presented in Section~\ref{sec:6:NumericalExperiments}, with more details in the Appendix.

\TODOL{Rename GLR, GLRT and GLR Bernoulli to B-GLRT like what we did in our GRETSI 2019 paper \cite{Besson2019Gretsi}.}

% -----------------------------------------------------------------
\section{The Piece-Wise Stationary Bandit Model}
\label{sec:6:BanditSetting}

@@

A policy $\pi$ chooses the next arm to play based on the sequence of past plays and obtained rewards.
The performance of $\pi$ is measured by its (piece-wise stationary) \emph{regret}, the difference between the expected reward obtained by an oracle policy playing an optimal arm $i^*_t$ at time $t$, and that of the policy $\pi$:
%
\begin{equation}\label{def:6:nti}
    R_T^{\pi} = \E\left[\sum_{t=1}^T \left(\mu_{i^*_t}(t) - \mu_{I_t}(t)\right)\right].
\end{equation}

In the piece-wise \iid{} model, we furthermore assume that there is a (relatively small) number of \emph{breakpoints}, denoted by $\Upsilon_T:=\sum_{t=1}^{T-1} \indic\left(\exists i\in\{1,\dots,K\}: \mu_t(i) \neq \mu_{t+1}(i)\right)$.
We define the $k$-th breakpoint by
\begin{equation}
    \tau^{(k)} = \inf\{t > \tau^{(k-1)} : \exists i : \mu_i(t) \neq \mu_{i}(t+1)\}.
\end{equation}
%
Hence for $t\in[\tau^{(k)} + 1,\tau^{(k+1)}]$, the rewards $(X_{i,t})$ associated to each arms are \iid.

Note than when a breakpoint occurs, we do not assume that all the arms means  change, but that \emph{there exists} an arm whose mean has changed. Depending on the application, many scenario can be meaningful: changes occurring on all arms simultaneously (due to some exogenous event), or only a few arms change at each breakpoint. Introducing the number of change-points  on arm $i$, defined as $\NCi:=\sum_{t=1}^{T-1} \indic\left(\mu_t(i) \neq \mu_{t+1}(i)\right)$, it clearly holds that
$\NCi \leq \Upsilon_T$, but there can be an arbitrary difference between these two quantities for some arms. Letting $C_T := \sum_{i=1}^K \NCi$ be the total number of change-points on the arms, one can have $C_T \in \{ \Upsilon_T, \dots, K\Upsilon_T \}$.

\paragraph{An interesting generalization}
%
The piece-wise stationary bandit model can be viewed as an interpolation between stationary and adversarial models, as the stationary model corresponds to $\Upsilon_T = 0$, while the adversarial model can be considered as a special (worst) case, with $\Upsilon_T = T$. However, analyzing an algorithm for the piece-wise stationary bandit model requires to assume a small number of changes, typically $\Upsilon_T = o(\sqrt{T})$.

\TODOL{Speak about the article \cite{WeiSrivastava18} and their point of view on $\Upsilon_T = \bigO{T^{\nu}}$ with known $\nu$ but unknown $\Upsilon_T$}

% -----------------------------------------------------------
\section{Related Work}
\label{sec:6:relatedWork}

We review previous works that studied the piece-wise stationary bandit model, or variants of this model.
To the authors knowledge, all the previous works are based on the idea of combining a classical bandit policy, such as Thompson sampling, \UCB{} or EXP3, with a strategy to account for changes in arms' distributions.
Following the vocabulary used in previous works, we make the distinction between passively and actively adaptive strategies.
Active strategies monitors the rewards of each arm by using a change detection algorithm (\cite{Basseville93}), and reset the history of pulls and rewards of one or all the arms as soon as a change is detected.
Passive strategies are tuned, with a discount factor or a limited memory size, while active strategies are usually tuned with a threshold for the statistical test.

@@

Both D-UCB and SW-UCB builds on a stationary policy,
but for example EXP3.S from \cite{Auer02NonStochastic}
builds on the EXP3 policy, which is designed for the adversarial case,
and achieves a larger regret upper-bound.
We also found in our experiments that it always performed worse that the reference stationary algorithms (\eg, \klUCB), so we do not include it in Section~\ref{sec:6:NumericalExperiments}.
Similarly, previous works showed that older algorithms have no or weaker regret guarantees, and have been proved to be less efficient empirically, or are designed for more specific settings.
%
For instance, Windowed-Mean Shift from \cite{YuMannor09} is generic and combines any efficient bandit policy with a CD test based on a sliding window, but there approach is not applicable to the bandit setting as they consider side observations.
%
A different setting where the quantity of interest is not $\Upsilon_T$ but $V_T$ the total variational budget, was introduced by \cite{Besbes14stochastic}, for which they proposed the RExp3 algorithm.
%
The RExp3 algorithm (\cite{Besbes14stochastic}) can also be qualified as passively adaptive: it is based on (non-adaptive) restarts of the EXP3 algorithm. Note that this algorithm is introduced for a different setting, where the quantity of interest is not $\Upsilon_T$ but a quantity $V_T$ called the total variational budget (satisfying $\Delta^{\text{change}} \Upsilon_T \leq V_T \leq \Upsilon_T$ with $\Delta^{\text{change}}$ the minimum magnitude of a change-point). A  $\cO(V_T^{1/3} T^{2/3})$ regret bound is proved, which is weaker than existing results in our setting. Hence we do no include this algorithm in our experiments.
%

The idea of using a simple discount factor, as for D-UCB, was recently adapted to a Bayesian policy, with the Discounted Thompson sampling (DTS) algorithm presented by \cite{RajKalyani17}.
Even if no theoretical guarantee was given, it can be empirically very efficient, but we found that DTS is not robust in the choice of if its discount factor $\gamma\in(0,1)$, contrarily to what was highlighted in their paper.
The DTS algorithm can perform well in practice, for instance with $\gamma=0.75$ on our Problems $1$ and $2$ (see Section~\ref{sec:6:NumericalExperiments}). However, no theoretical guarantees are given for this strategy, and our experiments did not really confirm the robustness to $\gamma$.
%
In general, we found that passively adaptive approaches can be efficient when their parameters are well tuned, but our experiments show that actively adaptive algorithms perform significantly better.

@@

When all their parameters are tuned correctly, both approaches are proved to achieve a good regret upper-bound.
As both results are valid under different assumptions, we consider the two results to be the current state-of-the-art.
As for previous works, tuning their parameters requires to know both the horizon $T$ and the number of breakpoints $\Upsilon_T$, but most importantly it requires a prior knowledge of the problem difficulty, by assuming to know a lower bound on both the time between changes (\eg, $L$ for \MUCB) and on the changes on the means of arms (\eg, $\varepsilon$ for \CUSUM).
%
\CUSUMUCB{} achieves a regret bounded by $\cO(\sqrt{\Upsilon_T T \log(T / \Upsilon_T)})$, but only for Bernoulli distributions,
and when its $4$ parameters ($\alpha,\varepsilon,M,h$) are tuned based on problem-dependent knowledge.
%
\MUCB{} achieves a regret bounded by $\cO(\sqrt{\Upsilon_T T \log(T)})$, for bounded distributions, and when its $4$ parameters ($\alpha,w,b,\gamma$) are also tuned based on a problem-dependent knowledge of $\tilde{\delta}$ and $L$ (see Assumption~1 and Remark~1).

\CUSUMUCB{} is based on a rather complicated two-sided \CUSUM{} test, that uses the first $M$ samples from one arm to compute an initial average, and then detects whether a drift of size larger than $\varepsilon$ occurred from this value by checking whether a random walk based on the remaining observations crosses a threshold $h$. Thus it requires the tuning of three parameters, $M$, $\varepsilon$ and $h$. \CUSUMUCB{} performs \emph{local restarts} using this test, to reset the history of \emph{one arm} for which the test detects a change.
\MUCB{} uses a much simpler test, based on the $w$ most recent observations from an arm: a change is detected if the absolute difference between the empirical means of the first and second halves of those $w$ observations exceeds a threshold $h$. So it requires the tuning of two parameters, $w$ and $h$. \MUCB{} performs \emph{global restarts} using this test, to reset the history of \emph{all arms} whenever the test detects a change on one of the arms.
Compared to \CUSUMUCB{}, note that \MUCB{} is numerically much simpler as it only uses a bounded memory, of order $\bigO{K w}$ for $K$ arms.

\paragraph{Expert aggregation.}
%
The second line of research of actively adaptive algorithms uses a Bayesian point of view.
A Bayesian CD algorithm is combined with Thompson Sampling in \cite{MellorShapiro13},
and more recently \cite{Alami17} introduced the Memory Bandit algorithm. It is presented as efficient empirically, but no theoretical guarantee are given, and due to its complexity, we do not include it in our experiments.
The idea behind Memory Bandit is to use an expert aggregation algorithm, like EXP4 from \cite{Auer02}, modified to efficiently aggregate an growing number of experts from techniques presented in \cite{Mourtada17}.
At each time step, a new expert is introduced, and experts corresponds to different Thompson sampling algorithms, each using a different history of pulls and rewards. Intuitively, after a change-point the newest experts will soon become the most efficient, as they are learning by using rewards drawn from the new distribution(s).
%
Note that the regret bound for these methods are still open-problems,
therefore we choose to not include them in our experiments.

\paragraph{Advantages of our approach}
%
\CUSUMUCB{} and \MUCB{} are both analyzed under some reasonable assumptions on the problem parameters --the means $(\mu_i(t))$-- mostly saying that the breakpoints are sufficiently far away from each other.
%
However, the proposed guarantees only hold for parameters \emph{tuned using some prior knowledge of the means}.
Indeed, while in both cases the threshold $h$ can be set as a function of the horizon $T$ and the number of breakpoints  $\Upsilon_T$ (also needed by previous approaches to obtain the best possible bounds), the parameter $\varepsilon$ for \CUSUM{} and $w$ for \MUCB{} require the knowledge of $\Delta^{\text{change}}$ the smallest magnitude of a change-point.
%
In this chapter, we propose \emph{the first algorithm that does not require this knowledge}, and still attains a $\cO(\sqrt{\Upsilon_T T\log(T)})$ regret.
Moreover we propose the first comparison of the use of local and global restarts within an adaptive algorithm, by studying two variants of our algorithm.
This study is supported by both theoretical and empirical results.
%
Finally, on the practical side, while we can note that the proposed GLR test is more complex to implement than the test used by \MUCB, we propose two heuristics to speed it up while not losing  much in terms of regret.

@@

% -----------------------------------------------------------------
\section{The Bernoulli GLR Change-Point Detector}
\label{sec:6:ChangePointDetector}

Sequential change-point detection has been extensively studied in the statistical community (see, \eg, \cite{Basseville93} for a survey). In this section, we are interested in detecting changes on the mean of a probability distribution with bounded support. Assuming that  we collect independent samples $X_1,X_2,\ldots$ all from some distribution supported in $[0,1]$. We want to discriminate between two possible scenarios: all the samples come from distributions that have a common mean $\mu_0$, or there exists a \emph{change-point} $\tau \in\N^*$ such that $X_1,\ldots,X_\tau$ have some mean $\mu_0$ and $X_{\tau +1},X_{\tau+2},\ldots$ have a different mean $\mu_1 \neq \mu_0$.
%
A sequential change-point detector is a stopping time $\hat\tau$ with respect to the filtration $\cF_t = \sigma(X_1,\dots,X_t)$ such that $(\hat \tau < \infty)$ means that we reject the hypothesis
$\cH_0 : \left(\exists \mu_0 \in [0,1]: \forall i\in\N, \bE[X_i] = \mu_0\right)$.
